<!--
This is for including Chapter 1.  Notice that it's also good practice to name your chunk.  This will help you debug potential issues as you knit.  The chunk above is called intro and the one below is called chapter1.  Feel free to change the name of the Rmd file as you wish, but don't forget to change it here from chap1.Rmd.
-->

<!--
The {#rmd-basics} text after the chapter declaration will allow us to link throughout the document back to the beginning of Chapter 1.  These labels will automatically be generated (if not specified) by changing the spaces to hyphens and capital letters to lowercase.  Look for the reference to this label at the beginning of Chapter 2.
-->

# R Markdown Basics {#rmd-basics}

Here is a brief introduction into using _R Markdown_. _Markdown_ is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. _R Markdown_ provides the flexibility of _Markdown_ with the implementation of **R** input and output.  For more details on using _R Markdown_ see <https://rmarkdown.rstudio.com>.  

Be careful with your spacing in _Markdown_ documents.  While whitespace largely is ignored, it does at times give _Markdown_ signals as to how to proceed.  As a habit, try to keep everything left aligned whenever possible, especially as you type a new paragraph.  In other words, there is no need to indent basic text in the Rmd document (in fact, it might cause your text to do funny things if you do).

## Lists

It's easy to create a list.  It can be unordered like

* Item 1
* Item 2

or it can be ordered like

1. Item 1
4. Item 2

Notice that I intentionally mislabeled Item 2 as number 4.  _Markdown_ automatically figures this out!  You can put any numbers in the list and it will create the list.  Check it out below.

To create a sublist, just indent the values a bit (at least four spaces or a tab).  (Here's one case where indentation is key!)

1. Item 1
1. Item 2
1. Item 3
    - Item 3a
    - Item 3b

## Line breaks

Make sure to add white space between lines if you'd like to start a new paragraph.  Look at what happens below in the outputted document if you don't:

Here is the first sentence.  Here is another sentence.  Here is the last sentence to end the paragraph.
This should be a new paragraph.

*Now for the correct way:* 

Here is the first sentence.  Here is another sentence.  Here is the last sentence to end the paragraph.

This should be a new paragraph.

## R chunks

When you click the **Knit** button above a document will be generated that includes both content as well as the output of any embedded **R** code chunks within the document. You can embed an **R** code chunk like this (`cars` is a built-in **R** dataset):

```{r cars}
knitr::kable(summary(cars))
```

## Inline code

If you'd like to put the results of your analysis directly into your discussion, add inline code like this:

> The `cos` of $2 \pi$ is `r cos(2*pi)`. 

Another example would be the direct calculation of the standard deviation:

> The standard deviation of `speed` in `cars` is `r sd(cars$speed)`.

One last neat feature is the use of the `ifelse` conditional statement which can be used to output text depending on the result of an **R** calculation:

> `r ifelse(sd(cars$speed) < 6, "The standard deviation is less than 6.", "The standard deviation is equal to or greater than 6.")`

Note the use of `>` here, which signifies a quotation environment that will be indented.

As you see with `$2 \pi$` above, mathematics can be added by surrounding the mathematical text with dollar signs.  More examples of this are in [Mathematics and Science] if you uncomment the code in [Math].  

## Including plots

You can also embed plots. For example, here is a way to use the base **R** graphics package to produce a plot using the built-in `pressure` dataset:



Note that the `echo=FALSE` parameter was added to the code chunk to prevent printing of the **R** code that generated the plot. There are plenty of other ways to add chunk options (like `fig.height` and `fig.width` in the chunk above).  More information is available at <https://yihui.org/knitr/options/>.  

Another useful chunk option is the setting of `cache=TRUE` as you see here.  If document rendering becomes time consuming due to long computations or plots that are expensive to generate you can use knitr caching to improve performance.  Later in this file, you'll see a way to reference plots created in **R** or external figures.

## Loading and exploring dinosaur diet data

This analysis explores feeding records of various dinosaur species, including their diet types, food preferences, and amount of food consumed across multiple days.

```{r loaddata}
# Read the CSV data
dino_data <- read.csv('data/dino_data.csv')

# Clean column names by removing spaces and making them R-friendly
names(dino_data) <- c("Day", "Time", "Species", "Diet", "Food_Type", "Kg_Food")
```

Let's examine the basic structure and summary of our dataset:

```{r datasummary}
dim(dino_data)
knitr::kable(summary(dino_data[1:4]))
```

Loading required packages for analysis:

```{r loadpkgs, message=FALSE}
pkg <- c("dplyr", "ggplot2", "knitr")
new.pkg <- pkg[!(pkg %in% installed.packages())]
if (length(new.pkg)) {
  install.packages(new.pkg, repos = "https://cran.rstudio.com")
}
library(dplyr)
library(ggplot2)
library(knitr)
```

```{r setup, include=FALSE}
# Set default figure size and other knitr options
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 4,
  out.width = "80%",
  fig.align = 'center'
)
```

Let's analyze feeding patterns by species:

```{r feedinganalysis}
# Calculate average daily consumption by species
avg_consumption <- dino_data %>%
  group_by(Species) %>%
  summarize(
    avg_daily_kg = mean(Kg_Food),
    total_kg = sum(Kg_Food),
    diet_type = first(Diet),
    n_meals = n()
  ) %>%
  arrange(desc(avg_daily_kg))
# Simple table output
kable(avg_consumption,
      caption = "Average consumption",
      col.names = c("Species", "Avg Kg per Meal", "Total Kg Eaten", 
                    "Diet Type", "Number of Meals"))
```

Analyzing diet types:

```{r dietanalysis}
diet_summary <- dino_data %>%
  group_by(Diet) %>%
  summarize(
    avg_meal_size = mean(Kg_Food),
    max_meal = max(Kg_Food),
    total_consumption = sum(Kg_Food),
    n_meals = n()
  )

kable(diet_summary,
      caption = "Diet summary",
      col.names = c("Diet Type", "Avg Meal Size (kg)", 
                    "Max Meal Size (kg)", "Total Food (kg)", "Number of Meals"))
```

Visualizing feeding patterns:

```{r dailypattern, fig.height=6, fig.width=10, fig.cap="Feeding Patterns Throughout the Day by Diet Type"}
ggplot(dino_data, aes(x = Time, y = Kg_Food, color = Diet)) +
  geom_point(size = 3, alpha = 0.7) +
  facet_wrap(~Diet) +
  theme_minimal() +
  labs(title = "Feeding Patterns by Diet Type",
       x = "Time of Day",
       y = "Food Consumed (kg)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Distribution of meal sizes:

```{r mealdistribution, fig.height=6, fig.width=8, fig.cap="Distribution of Meal Sizes by Diet Type"}
ggplot(dino_data, aes(x = Diet, y = Kg_Food, fill = Diet)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
  theme_minimal() +
  labs(title = "Meal Sizes by Diet Type",
       x = "Diet Category",
       y = "Food Consumed (kg)") +
  theme(legend.position = "none")
```

Food type preferences:

```{r foodpreferences}
food_preferences <- dino_data %>%
  group_by(Diet, Food_Type) %>%
  summarize(
    times_eaten = n(),
    avg_portion = mean(Kg_Food)
  ) %>%
  arrange(Diet, desc(times_eaten))

kable(food_preferences,
      caption = "Food preferences",
      col.names = c("Diet Type", "Food Type", 
                    "Times Eaten", "Average Portion (kg)"))
```

Daily consumption trends:

```{r dailyconsumption, fig.height=6, fig.width=10, fig.cap="Daily Food Consumption Trends by Diet Type"}
daily_total <- dino_data %>%
  group_by(Day, Diet) %>%
  summarize(total_food = sum(Kg_Food))

ggplot(daily_total, aes(x = Day, y = total_food, color = Diet, group = Diet)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Daily Consumption by Diet Type",
       x = "Day",
       y = "Total Food Consumed (kg)")
```

## Additional resources

- _Markdown_ Cheatsheet - <https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet>

- _R Markdown_
    - Reference Guide - <https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf>
    - Cheatsheet - <https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf>

- _RStudio IDE_
    - Cheatsheet - <https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf>
    - Official website - <https://rstudio.com/products/rstudio/>

- Introduction to `dplyr` - <https://cran.rstudio.com/web/packages/dplyr/vignettes/dplyr.html>

- `ggplot2`
    - Documentation - <https://ggplot2.tidyverse.org/>
    - Cheatsheet - <https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf>

